---
title: "Skill Testing MOM6 Oxygen"
author: "Olivia Gemmell"
date: "27 Aug 2025"
format: html
toc: true
editor: visual
---

## Purpose

There is a lack of environmental data—particularly dissolved oxygen—at the same temporal and spatial resolution as biological data, which can limit species distribution analyses. Estimates from oceanographic models can be used in place of direct in situ data, but their ability to accurately represent in situ data remains untested. Here, we skill test 3D MOM6 oxygen data outputs against in situ [CTD data](https://www.fisheries.noaa.gov/west-coast/science-data/oceanographic-and-ecosystem-sampling-during-pacific-hake-survey#measuring-conductivity-temperature-and-depth) collected during the Joint U.S.-Canada Pacific Hake Acoustic Trawl Survey following two methods:

1.  A simple nearest neighbour analysis

2.  Methodology from Indivero et al., 2025

Using these methods, we evaluate the validity of MOM6 oxygen outputs for use in age-structured short-term forecasts of Pacific hake.

```{r setup,include=F,warning=F,message=F}
library(tidyverse)
library(rnaturalearth)
library(sf)
library(here)
library(ncdf4)
library(marmap)
library(terra)
library(tidync)
library(viridis)
library(readxl)
library(purrr)
library(mapview)
library(devtools)
library(FishStatsUtils)
library(readr)
library(sdmTMB)
library(ggpubr)
library(furrr)
library(Metrics)
library(patchwork)
library(RColorBrewer)
library(bestNormalize)
library(e1071)

# create plot theme
plot_theme <- theme_bw() +
          theme(panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.border = element_blank(),
            strip.background = element_blank(),
            strip.text = element_blank()
          ) +
          theme(axis.line = element_line(color = "black")) +
          theme(axis.text = element_text(size = 11)) +
          theme(axis.title = element_text(size = 12)) +
          theme(legend.text = element_text(size = 11))

```

## Prepare Data

### Process in situ data

First we process the in situ CTD data so it's ready to use to calculate RMSE. See 01-process_o2_obs.R for code.

```{r}
#| echo: false
in_situ <- readRDS(here::here("test_cases/mom6/data/all_o2_dat_region.rds"))
glimpse(in_situ)
```

### Create region polygons and add NOAA bathymetry

Next, we create region polygons and add NOAA bathymetry data for model fitting later. We create separate regions for British Columbia and the California Current but we will be focusing on the combined region that includes both areas (i.e., "bc_cc") for this anaylsis. We also restrict the dataset to depths less than 500 m because the CTD casts only go that deep. See 02-estimate_depths_by_region.R for code.

```{r}
#| echo: false

# load regional polygons
regions.hull <- readRDS(here::here("test_cases/mom6/data/regions_hull.rds"))
# Map regions
mapview(regions.hull)

```

### Prepare MOM6 data

Then we prepare the MOM6 data for analysis. Here we grab 3D MOM6 oxygen output from the [CEFI portal](https://psl.noaa.gov/cefi_portal/) by year, convert it to a dataframe and match the format to the in situ data (i.e., match column naming, convert o2 units etc.). See 03-prep_mom6.R for code.

## Nearest Neighbor Analysis

We start our exploration by conducting a simple nearest neighbor analysis to evaluate how well the raw MOM6 output matches the in situ data. MOM6 data are interpolated using bilinear interpolation between spatial point (lat, lon), and linear interpolation between depths.

### Initial data checks

```{r}
#| echo: false
#| warning: false

nn <- readRDS(here::here("test_cases/mom6/data/nearest_neighbor.rds"))
# still have some NAs here because I haven't coded filling in NA values for the 3D version of add_mom() yet
# I'm subsetting by the raw MOM6 values (o2_nn, includes bilinear interpolation)
# because the linear interpolation for depth at NA points was inprecise  
nn <- nn[!is.na(nn$o2_nn),]

# Find the global O2 range across both columns
o2_range <- range(c(nn$o2, nn$o2_nn_interp), na.rm = TRUE)

# histogram of in situ data
p1 <- ggplot(nn, aes(x = o2)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_viridis_d() +
  scale_x_continuous(limits = o2_range) + 
  scale_y_continuous(limits = c(0, 10000)) + 
  plot_theme +
  labs(
    title = "In situ",
    x = "Dissolved Oxygen (µmol/kg)",
    y = "Count"
  )

# histograms of MOM6 data
p2 <- ggplot(nn, aes(x = o2_nn_interp)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_viridis_d() +
  scale_x_continuous(limits = o2_range) + 
  scale_y_continuous(limits = c(0, 10000)) + 
  plot_theme +
  labs(
    title = "MOM6",
    x = "Dissolved Oxygen (µmol/kg)",
    y = "Count"
  )


# Combine plots side by side with shared legend
p1 + p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")


```

```{r}
#| echo: false

# create 1:1 plot of predicted and observed
ggplot(data = nn, aes(x = o2, y = o2_nn_interp, col = depth)) +
          geom_point() +
          scale_colour_gradientn(colours = brewer.pal(9, "Blues")[3:9],  
          name = "Depth") +
          theme(legend.position = "bottom") +
          plot_theme +
          labs(x = "In situ (CTD)", y = "MOM6 o2") +
          geom_abline(intercept = 0, slope = 1)+
          theme(legend.position="bottom")

```

```{r}
#| echo: false

# calculate and plot residuals
nn$resids <- nn$o2 - nn$o2_nn_interp

# plot
ggplot(data = nn, aes(x = (o2_nn_interp), y = resids, col = depth)) +
    geom_point() +
    labs(x = "MOM6 o2", y = "Residuals (in situ - MOM6") +
    plot_theme +
    scale_colour_gradientn(colours = brewer.pal(9, "Blues")[3:9],  
          name = "Depth") +
  theme(legend.position = "bottom") 

```

```{r}
#| echo: false

# plot oxygen by depth
ggplot(data = nn, aes(x = (o2_nn_interp), y = depth)) +
    geom_point(alpha = 0.1) +
    labs(x = "MOM6 o2", y = "Depth") +
    scale_y_reverse() +
    plot_theme +
    theme(legend.position = "none") 

```

```{r}
#| echo: false
#| message: false
#| warning: false

# load coastline data
load(here::here("test_cases/mom6/data/namerica_state_wgs_df.RData"))

# calculate mean o2 by location- in situ
o2_by_location <- nn %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2, na.rm = TRUE)) %>%
  ungroup()

# calculate mean o2 by location- MOM6
o2_by_location_mom6 <- nn %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2_nn_interp, na.rm = TRUE)) %>%
  ungroup()

# Merge in situ and MOM6 mean O2 by location
o2_diff <- o2_by_location %>%
  rename(mean_o2_obs = mean_o2) %>%
  inner_join(o2_by_location_mom6 %>% rename(mean_o2_mom6 = mean_o2),
             by = c("longitude", "latitude")) %>%
  mutate(diff_o2 = mean_o2_obs - mean_o2_mom6)

# Find the global O2 range across both datasets
o2_range <- range(c(o2_by_location$mean_o2, o2_by_location_mom6$mean_o2), na.rm = TRUE)

# Find symmetric range for difference plot color scale
diff_range <- max(abs(o2_diff$diff_o2), na.rm = TRUE)

# First plot: In situ O2
p1 <- ggplot(o2_by_location) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean In Situ O2 Values") +
  scale_color_viridis(option = "viridis", limits = o2_range, name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Second plot: MOM6 O2
p2 <- ggplot(o2_by_location_mom6) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean MOM6 O2 Values") +
  scale_color_viridis(option = "viridis", limits = o2_range, name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Third plot: Difference (MOM6 - In Situ)
p3 <- ggplot(o2_diff) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = diff_o2),
             size = 0.5, alpha = 0.7) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Difference: In Situ - MOM6", color = "ΔO2 (µmol/kg)") +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-diff_range, diff_range)
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10))

# Combine plots side by side with shared legend
p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")


```

### RMSE results

#### All data

```{r}
#| echo: false

# Calculate RMSE by year
nn %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, o2_nn_interp)
  )

# Calculate total RMSE 
rmse(nn$o2, nn$o2_nn_interp)
```

#### Hake upper tolerance (o2 \< 119 µm/kg)

Let's narrow the range of values to include only o2 values where hake start showing altered behavior or distribution due to low oxygen (Moriarty et al., 2020). In theory, large oxygen values need to be less precise because they won't influence distribution (i.e., we only need precise o2 estimates at lower values when oxygen influences distribution).

```{r}
#| echo: false

### look at hypoxic zones (Barth et al., 2024) ###
nn_hypoxic <- nn[nn$o2 <= 119, ]
# Calculate RMSE by year
nn_hypoxic %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, o2_nn_interp)
  )
# Calculate total RMSE 
rmse(nn_hypoxic$o2, nn_hypoxic$o2_nn_interp)

```

#### Hake habitat (100-500 m)

Next let's see how well oxygen values are predicted within hake habitat.

```{r}
#| echo: false

### look at hake habitat ###
nn_habitat <- nn[nn$depth > 100 & nn$depth < 500, ]
# Calculate RMSE by year
nn_habitat %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, o2_nn_interp)
  )
# Calculate total RMSE 
rmse(nn_habitat$o2, nn_habitat$o2_nn_interp)

```

#### Upper tolerance within hake habitat (o2 \< 119 µm/kg between 100-500 m)

And finally, let's see how well o2 values within the upper tolerance are estimated in hake habitat.

```{r}
#| echo: false

### look at hypoxic zones in hake habitat ###
nn_hh <- nn_habitat[nn_habitat$o2 <= 119, ]
# Calculate RMSE by year
nn_hh %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, o2_nn_interp)
  )
# Calculate total RMSE 
rmse(nn_hh$o2, nn_hh$o2_nn_interp)

```

## Skill Testing Analysis

Then we conduct an analysis following methodology from Indivero et al., 2025 to improve the skill of the MOM6 output. Here, we extract MOM6 values at all vertical levels shallower than 500 m within a specified region (e.g., bc_cc, bc, or cc), fit a spatial model to the extracted MOM6 data, predict back to the location and depth of the in situ data, and calculate RMSE to evaluate skill.

### Initial data checks

```{r}
#| echo: false

# load(here::here("test_cases/mom6/output/o2_models/bc_cc.RData"))
# region_dat2 <- output[[1]]$mom6_data

region_dat <- read_rds(file = here::here("test_cases/mom6/data/mom6/mom6_bc_cc.rds"))
# restrict to < 500 m (hake habitat)
region_dat <- region_dat[region_dat$depth <= 500, ]
# remove negative values and NAs
region_dat <- region_dat[!is.na(region_dat$o2 >= 0), ]
# add lat and lon columns
# Convert back to sf object using the geometry column
region_sf <- st_as_sf(region_dat)
# Extract coordinates
coords <- st_coordinates(region_sf)
# Add longitude and latitude to original dataframe
region_dat$longitude <- coords[, 1]
region_dat$latitude  <- coords[, 2]

# Plot MOM6 depth by o2
ggplot(data = region_dat, aes(x = o2, y = depth)) +
    geom_point(alpha = 0.1) +
    labs(x = "MOM6 o2", y = "Depth") +
    scale_y_reverse() +
    plot_theme +
    theme(legend.position = "none") 

```

```{r}
#| echo: false
#| message: false
#| warning: false

# calculate mean o2 by location- MOM6
o2_by_location_mom6 <- region_dat %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2, na.rm = TRUE)) %>%
  ungroup()

# Plot MOM6 O2
ggplot(o2_by_location_mom6) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(region_dat$longitude, na.rm = TRUE), max(region_dat$longitude, na.rm = TRUE)) +
  ylim(min(region_dat$latitude, na.rm = TRUE), max(region_dat$latitude, na.rm = TRUE)) +
  labs(title = "Mean MOM6 O2 Values") +
  scale_color_viridis(option = "viridis", name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))
```

```{r}
#| echo: false
#| warning: false

# Find the global O2 range across both columns
o2_range <- range(c(nn$o2, region_dat$o2), na.rm = TRUE)

# histogram of in situ data
p1 <- ggplot(nn, aes(x = o2)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_viridis_d() +
  scale_x_continuous(limits = o2_range) + 
  # scale_y_continuous(limits = c(0, 10000)) + 
  plot_theme +
  labs(
    title = "In situ",
    x = "Dissolved Oxygen (µmol/kg)",
    y = "Count"
  )

# histograms of MOM6 data
p2 <- ggplot(region_dat, aes(x = o2)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_viridis_d() +
  scale_x_continuous(limits = o2_range) +
  # scale_y_continuous(limits = c(0, 10000)) +
  plot_theme +
  labs(
    title = "MOM6",
    x = "Dissolved Oxygen (µmol/kg)",
    y = "Count"
  )


# Combine plots side by side with shared legend
p1 + p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")

```

The data is left skewed so let's try and make it look more normally distributed using a quantile transformation.

```{r}
#| echo: false
#| warning: false

# Apply ordered quantile normalization to o2 column
bn <- orderNorm(region_dat$o2)
# Create transformed column
region_dat$o2_transformed <- predict(bn)

# Recover original o2 values
region_dat$o2_original <- predict(bn, newdata = region_dat$o2_transformed, inverse = TRUE)

# histograms of MOM6 data
ggplot(region_dat, aes(x = o2_transformed)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_viridis_d() +
  plot_theme +
  labs(
    title = "MOM6",
    x = "Dissolved Oxygen (µmol/kg)",
    y = "Count"
  )
```

### Fit GLMM to MOM6 output and calculate RMSE

Model structures tested:

### RMSE results

#### All data

```{r}
#| echo: false

# RMSE
rmse_year <- read_rds(here::here("test_cases/mom6/output/mom6_summary_bc_cc.rds"))
rmse_year

rmse <- read_rds(here::here("test_cases/mom6/output/mom6_rmsetotal_bc_cc.rds"))
rmse

```

#### Hake upper tolerance (o2 \< 119 µm/kg)

Again, we narrow the range of values to include only o2 values where hake start showing altered behavior or distribution due to low oxygen (Moriarty et al., 2020)...

```{r}
#| echo: false

load(here::here("test_cases/mom6/output/o2_models/bc_cc.RData"))
# combine all years of predictions
skill_preds <- rbind(output[[1]]$predictions,
                     output[[2]]$predictions,
                     output[[3]]$predictions,
                     output[[4]]$predictions)

### look at hypoxic zones ###
skill_hypoxic <- skill_preds[skill_preds$o2 <= 119, ]
# Calculate RMSE by year
skill_hypoxic %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, est)
  )
# Calculate total RMSE 
rmse(skill_hypoxic$o2, skill_hypoxic$est)
```

#### Hake habitat (100-500 m)

...and see how well oxygen values are predicted within hake habitat...

```{r}
#| echo: false

### look at hypoxic zones ###
skill_habitat <- skill_preds[skill_preds$depth > 100 & skill_preds$depth < 500, ]
# Calculate RMSE by year
skill_habitat %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, est)
  )
# Calculate total RMSE 
rmse(skill_habitat$o2, skill_habitat$est)
```

#### Upper tolerance within hake habitat (o2 \< 119 µm/kg between 100-500 m)

...and finally, let's see how well o2 values within the upper tolerance are estimated in hake habitat.

```{r}
#| echo: false

skill_hh <- skill_habitat[skill_habitat$o2 <= 119, ]
# Calculate RMSE by year
skill_hh %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, est)
  )
# Calculate total RMSE 
rmse(skill_hh$o2, skill_hh$est)

```

### 2011 Results

```{r}
#| echo: false

dat <- output[[1]]$predictions

# Plot depth by estimated o2
ggplot(data = dat, aes(x = est, y = depth)) +
    geom_point(alpha = 0.1) +
    labs(x = "Predicted o2", y = "Depth") +
    scale_y_reverse() +
    plot_theme +
    theme(legend.position = "none") 
```

```{r}
#| echo: false
#| message: false
#| warning: false

# calculate mean o2 by location- in situ
o2_by_location <- dat %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2, na.rm = TRUE)) %>%
  ungroup()

# calculate mean o2 by location- MOM6 model estimates
o2_by_location_mom6 <- dat %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(est, na.rm = TRUE)) %>%
  ungroup()

# Merge in situ and MOM6 mean O2 by location
o2_diff <- o2_by_location %>%
  rename(mean_o2_obs = mean_o2) %>%
  inner_join(o2_by_location_mom6 %>% rename(mean_o2_mom6 = mean_o2),
             by = c("longitude", "latitude")) %>%
  mutate(diff_o2 = mean_o2_obs - mean_o2_mom6)

# Find the global O2 range across both datasets
o2_range <- range(c(o2_by_location$mean_o2, o2_by_location_mom6$mean_o2), na.rm = TRUE)

# Find symmetric range for difference plot color scale
diff_range <- max(abs(o2_diff$diff_o2), na.rm = TRUE)

# First plot: In situ O2
p1 <- ggplot(o2_by_location) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean In Situ O2 Values") +
  scale_color_viridis(option = "viridis", limits = o2_range, name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Second plot: MOM6 O2
p2 <- ggplot(o2_by_location_mom6) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean predicted O2 Values") +
  scale_color_viridis(option = "viridis", limits = o2_range, name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Third plot: Difference (in situ - predicted)
p3 <- ggplot(o2_diff) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = diff_o2),
             size = 0.5, alpha = 0.7) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Difference: In Situ - predicted", color = "ΔO2 (µmol/kg)") +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-diff_range, diff_range)
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10))

# Combine plots side by side with shared legend
p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")

```

## Conclusions

The nearest neighbor analysis performed slightly better for all the data combined (RMSE = 37.61 vs. 41.62) but slightly worse when we narrowed the RMSE calculations to include only oxygen values that can influence hake distribution within hake habitat (RMSE = 25.75 vs. 23.35). RMSE values of 15-20 are considered within an acceptable range (Indivero pers. communication, 2025-08-22).

-   shallower areas have high oxygen values so less important to be precise

-   want more precision within hake habitat, specifically for our model- 250-300 m
