---
title: "Skill Testing MOM6 Oxygen"
author: "Olivia Gemmell"
date: "27 Aug 2025"
format: html
references:
  - id: indivero2025
    type: article-journal
    author:
      - family: Indivero
        given: Julia
      - family: Anderson
        given: Sean C.
      - family: Barnett
        given: Lewis A. K.
      - family: Pohl
        given: John E.
      - family: Rohan
        given: Sean K.
      - family: Siedlecki
        given: Samantha
      - family: Ward
        given: Eric J.
      - family: Essington
        given: Timothy E.
    issued:
      year: 2025
    title: "Skill Testing Oxygen Data for Distribution Modeling of Marine Species"
    container-title: "Fisheries Oceanography"
    DOI: "10.1111/fog.70005"
toc: true
editor: visual
---

## Purpose

There is a lack of environmental data—particularly dissolved oxygen—at the same temporal and spatial resolution as biological data, which can limit species distribution analyses. Estimates from oceanographic models can be used in place of direct in situ data, but their ability to accurately represent in situ data remains untested. Here, we skill test 3D MOM6 oxygen data outputs against in situ [CTD data](https://www.fisheries.noaa.gov/west-coast/science-data/oceanographic-and-ecosystem-sampling-during-pacific-hake-survey#measuring-conductivity-temperature-and-depth) collected during the Joint U.S.-Canada Pacific Hake Acoustic Trawl Survey following three methods:

1.  A nearest neighbour analysis

2.  Methodology from @indivero2025

3.  A binary confusion matrix analysis

Using these methods, we evaluate the validity of MOM6 oxygen outputs for use in age-structured short-term forecasts of Pacific hake.

```{r setup,include=F,warning=F,message=F}
library(tidyverse)
library(rnaturalearth)
library(sf)
library(here)
library(ncdf4)
library(marmap)
library(terra)
library(tidync)
library(viridis)
library(readxl)
library(purrr)
library(mapview)
library(devtools)
library(FishStatsUtils)
library(readr)
library(sdmTMB)
library(ggpubr)
library(furrr)
library(Metrics)
library(patchwork)
library(RColorBrewer)
library(bestNormalize)
library(e1071)

# create plot theme
plot_theme <- theme_bw() +
          theme(panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.border = element_blank(),
            strip.background = element_blank(),
            strip.text = element_blank()
          ) +
          theme(axis.line = element_line(color = "black")) +
          theme(axis.text = element_text(size = 11)) +
          theme(axis.title = element_text(size = 12)) +
          theme(legend.text = element_text(size = 11))

```

## Prepare Data

### Process in situ data

First we process the in situ CTD data so it's ready to use to calculate RMSE. See 01-process_o2_obs.R for code.

```{r}
#| echo: false
in_situ <- readRDS(here::here("test_cases/mom6/data/all_o2_dat_region.rds"))
glimpse(in_situ)
```

### Create region polygons and add NOAA bathymetry

Next, we create region polygons and add NOAA bathymetry data for model fitting later. We create separate regions for British Columbia and the California Current but we will be focusing on the combined region that includes both areas (i.e., "bc_cc") for this anaylsis. We also restrict the dataset to depths less than 500 m because the CTD casts only go that deep. See 02-estimate_depths_by_region.R for code.

```{r}
#| echo: false

# load regional polygons
regions.hull <- readRDS(here::here("test_cases/mom6/data/regions_hull.rds"))
# Map regions
mapview(regions.hull)

```

### Prepare MOM6 data

Then we prepare the MOM6 data for analysis. Here we grab 3D MOM6 oxygen output from the [CEFI portal](https://psl.noaa.gov/cefi_portal/) by year, convert it to a dataframe and match the format to the in situ data (i.e., match column naming, convert o2 units etc.). See 03-prep_mom6.R for code.

## Nearest Neighbor Analysis

We start our exploration by conducting a simple nearest neighbor analysis to evaluate how well the raw MOM6 output matches the in situ data. MOM6 data are interpolated using bilinear interpolation between spatial point (lat, lon), and linear interpolation between depths.

### Initial data checks

```{r}
#| echo: false
#| warning: false

nn <- readRDS(here::here("test_cases/mom6/data/nearest_neighbor.rds"))
# still have some NAs here because I haven't coded filling in NA values for the 3D version of add_mom() yet
# I'm subsetting by the raw MOM6 values (o2_nn, includes bilinear interpolation)
# because the linear interpolation for depth at NA points was inprecise  
nn <- nn[!is.na(nn$o2_nn),]

# Find the global O2 range across both columns
o2_range <- range(c(nn$o2, nn$o2_nn_interp), na.rm = TRUE)

# histogram of in situ data
p1 <- ggplot(nn, aes(x = o2)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_viridis_d() +
  scale_x_continuous(limits = o2_range) + 
  scale_y_continuous(limits = c(0, 10000)) + 
  plot_theme +
  labs(
    title = "In situ",
    x = "Dissolved Oxygen (µmol/kg)",
    y = "Count"
  )

# histograms of MOM6 data
p2 <- ggplot(nn, aes(x = o2_nn_interp)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_viridis_d() +
  scale_x_continuous(limits = o2_range) + 
  scale_y_continuous(limits = c(0, 10000)) + 
  plot_theme +
  labs(
    title = "MOM6",
    x = "Dissolved Oxygen (µmol/kg)",
    y = "Count"
  )


# Combine plots side by side with shared legend
p1 + p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")


```

The data distributions look similar.

```{r}
#| echo: false

# create 1:1 plot of predicted and observed
ggplot(data = nn, aes(x = o2, y = o2_nn_interp, col = depth)) +
          geom_point() +
          scale_colour_gradientn(colours = brewer.pal(9, "Blues")[3:9],  
          name = "Depth",
          breaks = scales::pretty_breaks(n = 3)) +
          theme(legend.position = "bottom") +
          plot_theme +
          labs(x = "In situ (CTD)", y = "MOM6 O2") +
          geom_abline(intercept = 0, slope = 1)+
          theme(legend.position="bottom")

```

```{r}
#| echo: false

# calculate and plot residuals
nn$resids <- nn$o2 - nn$o2_nn_interp

# plot oxygen and residuals
ggplot(data = nn, aes(x = (o2_nn_interp), y = resids, col = depth)) +
    geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
    labs(x = "MOM6 O2", y = "Residuals (in situ - MOM6)") +
    plot_theme +
    scale_colour_gradientn(colours = brewer.pal(9, "Blues")[3:9],
          name = "Depth",
          breaks = scales::pretty_breaks(n = 3)) +
  theme(legend.position = "bottom") 

```

```{r}
#| echo: false

# plot depth and residuals
ggplot(data = nn, aes(x = depth, y = resids, col = o2_nn_interp)) +
    geom_point() +
    labs(x = "Depth (m)", y = "Residuals (in situ - MOM6") +
    plot_theme +
    scale_colour_viridis(name = "MOM6 O2") +
  theme(legend.position = "bottom") 
```

Overall, MOM6 tends to overestimate oxygen compared to in situ values with greater overestimation at higher oxygen values and shallower depths.

```{r}
#| echo: false

# plot oxygen by depth
ggplot(data = nn, aes(x = (o2_nn_interp), y = depth)) +
    geom_point(alpha = 0.1) +
    labs(x = "MOM6 O2", y = "Depth") +
    scale_y_reverse() +
    plot_theme +
    theme(legend.position = "none") 

```

The extracted MOM6 values decrease with increasing depth, following the expected oxygen and depth relationship.

```{r}
#| echo: false
#| message: false
#| warning: false

# load coastline data
load(here::here("test_cases/mom6/data/namerica_state_wgs_df.RData"))

# calculate mean o2 by location- in situ
o2_by_location <- nn %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2, na.rm = TRUE)) %>%
  ungroup()

# calculate mean o2 by location- MOM6
o2_by_location_mom6 <- nn %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2_nn_interp, na.rm = TRUE)) %>%
  ungroup()

# Merge in situ and MOM6 mean O2 by location
o2_diff <- o2_by_location %>%
  rename(mean_o2_obs = mean_o2) %>%
  inner_join(o2_by_location_mom6 %>% rename(mean_o2_mom6 = mean_o2),
             by = c("longitude", "latitude")) %>%
  mutate(diff_o2 = mean_o2_obs - mean_o2_mom6)

# Find the global O2 range across both datasets- use these values for all following
# plots to keep them on the same scale
o2_range <- range(c(o2_by_location$mean_o2, o2_by_location_mom6$mean_o2), na.rm = TRUE)

# Find symmetric range for difference plot color scale- use these values for all 
# following plots to keep them on the same scale
diff_range <- max(abs(o2_diff$diff_o2), na.rm = TRUE)

# First plot: In situ O2
p1 <- ggplot(o2_by_location) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean In Situ O2 Values") +
  scale_color_viridis(option = "viridis", limits = c(0, max(o2_range)), name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Second plot: MOM6 O2
p2 <- ggplot(o2_by_location_mom6) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean MOM6 O2 Values") +
  scale_color_viridis(option = "viridis", limits = c(0, max(o2_range)), name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Third plot: Difference (MOM6 - In Situ)
p3 <- ggplot(o2_diff) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = diff_o2),
             size = 0.5, alpha = 0.7) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Difference: In Situ - MOM6", color = "ΔO2 (µmol/kg)") +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-diff_range, diff_range)
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10))

# Combine plots side by side with shared legend
p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")


```

Looking at mean (i.e., averaged over all depth layers) spatial differences, MOM6 tends to overestimate oxygen closer to the shoreline, where the influence of shallower, more oxygenated depth layers is greater.

### RMSE results

#### All data

Now let's look at the RMSE results, first looking at results using all the data combined.

```{r}
#| echo: false

# Calculate RMSE by year
nn %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, o2_nn_interp)
  )

# Calculate total RMSE 
rmse(nn$o2, nn$o2_nn_interp)
```

#### Hake oxygen preference boundary (o2 \< 119 µm/kg)

Let's narrow the range of values to include only o2 values where hake start showing altered behavior or distribution due to low oxygen (Moriarty et al., 2020). In theory, large oxygen values need to be less precise because they won't influence distribution (i.e., we only need precise O2 estimates at lower values when oxygen influences distribution).

```{r}
#| echo: false

### look at hypoxic zones (Barth et al., 2024) ###
nn_hypoxic <- nn[nn$o2 <= 119, ]
# Calculate RMSE by year
nn_hypoxic %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, o2_nn_interp)
  )
# Calculate total RMSE 
rmse(nn_hypoxic$o2, nn_hypoxic$o2_nn_interp)

```

#### Hake habitat (100-500 m)

Next let's see how well oxygen values are predicted within hake habitat.

```{r}
#| echo: false

### look at hake habitat ###
nn_habitat <- nn[nn$depth >= 100 & nn$depth <= 500, ]
# Calculate RMSE by year
nn_habitat %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, o2_nn_interp)
  )
# Calculate total RMSE 
rmse(nn_habitat$o2, nn_habitat$o2_nn_interp)

```

And look at the spatial differences of the data within hake habitat (i.e., excluding the mixed layer).

```{r}
#| echo: false
#| message: false
#| warning: false

# calculate mean o2 by location- in situ
o2_by_location <- nn_habitat %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2, na.rm = TRUE)) %>%
  ungroup()

# calculate mean o2 by location- MOM6
o2_by_location_mom6 <- nn_habitat %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2_nn_interp, na.rm = TRUE)) %>%
  ungroup()

# Merge in situ and MOM6 mean O2 by location
o2_diff <- o2_by_location %>%
  rename(mean_o2_obs = mean_o2) %>%
  inner_join(o2_by_location_mom6 %>% rename(mean_o2_mom6 = mean_o2),
             by = c("longitude", "latitude")) %>%
  mutate(diff_o2 = mean_o2_obs - mean_o2_mom6)

# Find the global O2 range across both datasets
# o2_range <- range(c(o2_by_location$mean_o2, o2_by_location_mom6$mean_o2), na.rm = TRUE)

# Find symmetric range for difference plot color scale
# diff_range <- max(abs(o2_diff$diff_o2), na.rm = TRUE)

# First plot: In situ O2
p1 <- ggplot(o2_by_location) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean In Situ O2 Values") +
  scale_color_viridis(option = "viridis", limits = c(0, max(o2_range)), name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Second plot: MOM6 O2
p2 <- ggplot(o2_by_location_mom6) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean MOM6 O2 Values") +
  scale_color_viridis(option = "viridis", limits = c(0, max(o2_range)), name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Third plot: Difference (MOM6 - In Situ)
p3 <- ggplot(o2_diff) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = diff_o2),
             size = 0.5, alpha = 0.7) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Difference: In Situ - MOM6", color = "ΔO2 (µmol/kg)") +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-diff_range, diff_range)
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10))

# Combine plots side by side with shared legend
p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")

```

#### Oxygen preference boundary within hake habitat (o2 \< 119 µm/kg between 100-500 m)

Then let's see how well O2 values within the oxygen preference boundary are estimated in hake habitat.

```{r}
#| echo: false

### look at hypoxic zones in hake habitat ###
nn_hh <- nn_habitat[nn_habitat$o2 <= 119, ]
# Calculate RMSE by year
nn_hh %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, o2_nn_interp)
  )
# Calculate total RMSE 
rmse(nn_hh$o2, nn_hh$o2_nn_interp)

```

#### Hake forecast work (250-300 m)

And finally, lets narrow the depth range further to look at the range of values we will focus on in our forecasting work.

```{r}
#| echo: false

### look at hake habitat ###
nn_forecast <- nn[nn$depth >= 250 & nn$depth <= 300, ]
# Calculate RMSE by year
nn_forecast %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, o2_nn_interp)
  )
# Calculate total RMSE 
rmse(nn_forecast$o2, nn_forecast$o2_nn_interp)

```

...and look at it spatially.

```{r}
#| echo: false
#| message: false
#| warning: false

# calculate mean o2 by location- in situ
o2_by_location <- nn_forecast %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2, na.rm = TRUE)) %>%
  ungroup()

# calculate mean o2 by location- MOM6
o2_by_location_mom6 <- nn_forecast %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2_nn_interp, na.rm = TRUE)) %>%
  ungroup()

# Merge in situ and MOM6 mean O2 by location
o2_diff <- o2_by_location %>%
  rename(mean_o2_obs = mean_o2) %>%
  inner_join(o2_by_location_mom6 %>% rename(mean_o2_mom6 = mean_o2),
             by = c("longitude", "latitude")) %>%
  mutate(diff_o2 = mean_o2_obs - mean_o2_mom6)

# First plot: In situ O2
p1 <- ggplot(o2_by_location) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean In Situ O2 Values") +
  scale_color_viridis(option = "viridis", limits = c(0, max(o2_range)), name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Second plot: MOM6 O2
p2 <- ggplot(o2_by_location_mom6) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean MOM6 O2 Values") +
  scale_color_viridis(option = "viridis", limits = c(0, max(o2_range)), name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Third plot: Difference (MOM6 - In Situ)
p3 <- ggplot(o2_diff) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = diff_o2),
             size = 0.5, alpha = 0.7) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Difference: In Situ - MOM6", color = "ΔO2 (µmol/kg)") +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-diff_range, diff_range)
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10))

# Combine plots side by side with shared legend
p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

## Indivero Analysis

Then we conduct an analysis following methodology from Indivero et al., 2025 to attempt to improve the skill of the MOM6 output. Here, we extract MOM6 values at all vertical levels shallower than 500 m within a specified region (e.g., bc_cc, bc, or cc), fit a spatial model to the extracted MOM6 data, predict back to the location and depth of the in situ data, and calculate RMSE to evaluate skill.

### Initial data checks before model fitting

```{r}
#| echo: false

# load(here::here("test_cases/mom6/output/o2_models/bc_cc.RData"))
# region_dat2 <- output[[1]]$mom6_data

region_dat <- read_rds(file = here::here("test_cases/mom6/data/mom6/mom6_bc_cc.rds"))
# restrict to < 500 m (hake habitat)
region_dat <- region_dat[region_dat$depth <= 500, ]
# remove negative values and NAs
region_dat <- region_dat[!is.na(region_dat$o2 >= 0), ]
# add lat and lon columns
# Convert back to sf object using the geometry column
region_sf <- st_as_sf(region_dat)
# Extract coordinates
coords <- st_coordinates(region_sf)
# Add longitude and latitude to original dataframe
region_dat$longitude <- coords[, 1]
region_dat$latitude  <- coords[, 2]

# Plot MOM6 depth by o2
ggplot(data = region_dat, aes(x = o2, y = depth)) +
    geom_point(alpha = 0.1) +
    labs(x = "MOM6 o2", y = "Depth") +
    scale_y_reverse() +
    plot_theme +
    theme(legend.position = "none") 

```

```{r}
#| echo: false
#| message: false
#| warning: false

# calculate mean o2 by location- MOM6
o2_by_location_mom6 <- region_dat %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2, na.rm = TRUE)) %>%
  ungroup()

# Plot MOM6 O2
ggplot(o2_by_location_mom6) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(region_dat$longitude, na.rm = TRUE), max(region_dat$longitude, na.rm = TRUE)) +
  ylim(min(region_dat$latitude, na.rm = TRUE), max(region_dat$latitude, na.rm = TRUE)) +
  labs(title = "Mean MOM6 O2 Values in 2011") +
  scale_color_viridis(option = "viridis", name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))
```

```{r}
#| echo: false
#| warning: false

# Find the global O2 range across both columns
o2_range_region <- range(c(nn$o2, region_dat$o2), na.rm = TRUE)

# histogram of in situ data
p1 <- ggplot(nn, aes(x = o2)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_viridis_d() +
  scale_x_continuous(limits = o2_range_region) + 
  # scale_y_continuous(limits = c(0, 10000)) + 
  plot_theme +
  labs(
    title = "In situ",
    x = "Dissolved Oxygen (µmol/kg)",
    y = "Count"
  )

# histograms of MOM6 data
p2 <- ggplot(region_dat, aes(x = o2)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_viridis_d() +
  scale_x_continuous(limits = o2_range_region) +
  # scale_y_continuous(limits = c(0, 10000)) +
  plot_theme +
  labs(
    title = "MOM6",
    x = "Dissolved Oxygen (µmol/kg)",
    y = "Count"
  )


# Combine plots side by side with shared legend
p1 + p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")

```

The data is left skewed but the Dharma residuals and model fit looked fine so we will use it as is (i.e., untransformed).

### Model structure

```{r}
#| echo: false


load(here::here("test_cases/mom6/output/o2_models/scaled_untransformed/bc_cc.RData"))
mod <- output[[1]]$model
mod

```

### RMSE results

#### All data

Now let's look at the RMSE results, first looking at results using all the data combined.

```{r}
#| echo: false

# RMSE
rmse_year <- read_rds(here::here("test_cases/mom6/output/scaled_untransformed/mom6_summary_bc_cc.rds"))
rmse_year

rmse <- read_rds(here::here("test_cases/mom6/output/scaled_untransformed/mom6_rmsetotal_bc_cc.rds"))
rmse

```

#### Hake oxygen preference boundary (o2 \< 119 µm/kg)

Again, we narrow the range of values to include only O2 values where hake start showing altered behavior or distribution due to low oxygen (Moriarty et al., 2020)...

```{r}
#| echo: false

# combine all years of predictions
skill_preds <- rbind(output[[1]]$predictions,
                     output[[2]]$predictions,
                     output[[3]]$predictions,
                     output[[4]]$predictions)

### look at hypoxic zones ###
skill_hypoxic <- skill_preds[skill_preds$o2 <= 119, ]
# Calculate RMSE by year
skill_hypoxic %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, est)
  )
# Calculate total RMSE 
rmse(skill_hypoxic$o2, skill_hypoxic$est)
```

#### Hake habitat (100-500 m)

...and see how well oxygen values are predicted within hake habitat...

```{r}
#| echo: false

### look at hypoxic zones ###
skill_habitat <- skill_preds[skill_preds$depth >= 100 & skill_preds$depth <= 500, ]
# Calculate RMSE by year
skill_habitat %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, est)
  )
# Calculate total RMSE 
rmse(skill_habitat$o2, skill_habitat$est)
```

...looking at spatial distributions...

```{r}
#| echo: false
#| message: false
#| warning: false

# calculate mean o2 by location- in situ
o2_by_location <- skill_habitat %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2, na.rm = TRUE)) %>%
  ungroup()

# calculate mean o2 by location- MOM6
o2_by_location_mom6 <- skill_habitat %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(est, na.rm = TRUE)) %>%
  ungroup()

# Merge in situ and MOM6 mean O2 by location
o2_diff <- o2_by_location %>%
  rename(mean_o2_obs = mean_o2) %>%
  inner_join(o2_by_location_mom6 %>% rename(mean_o2_mom6 = mean_o2),
             by = c("longitude", "latitude")) %>%
  mutate(diff_o2 = mean_o2_obs - mean_o2_mom6)

# First plot: In situ O2
p1 <- ggplot(o2_by_location) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean In Situ O2 Values") +
  scale_color_viridis(option = "viridis", limits = c(0, max(o2_range)), name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Second plot: MOM6 O2
p2 <- ggplot(o2_by_location_mom6) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean MOM6 O2 Values") +
  scale_color_viridis(option = "viridis", limits = c(0, max(o2_range)), name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Third plot: Difference (MOM6 - In Situ)
p3 <- ggplot(o2_diff) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = diff_o2),
             size = 0.5, alpha = 0.7) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Difference: In Situ - MOM6", color = "ΔO2 (µmol/kg)") +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-diff_range, diff_range)
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10))

# Combine plots side by side with shared legend
p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Oxygen preference boundary within hake habitat (o2 \< 119 µm/kg between 100-500 m)

... then looking at how well O2 values within the oxygen preference boundary are estimated in hake habitat...

```{r}
#| echo: false

skill_hh <- skill_habitat[skill_habitat$o2 <= 119, ]
# Calculate RMSE by year
skill_hh %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, est)
  )
# Calculate total RMSE 
rmse(skill_hh$o2, skill_hh$est)

```

#### Hake forecast work (250-300 m)

...and finally, narrowing the depth range further to look at the range of values we will focus on in our forecasting work...

```{r}
#| echo: false

### look at hake habitat ###
skill_forecast <- skill_preds[skill_preds$depth >= 250 & skill_preds$depth <= 300, ]
# Calculate RMSE by year
skill_forecast %>%
  group_by(year) %>%
  summarise(
    rmse = rmse(o2, est)
  )
# Calculate total RMSE 
rmse(skill_forecast$o2, skill_forecast$est)

```

...and looking at it spatially.

```{r}
#| echo: false
#| message: false
#| warning: false

# calculate mean o2 by location- in situ
o2_by_location <- skill_forecast %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2, na.rm = TRUE)) %>%
  ungroup()

# calculate mean o2 by location- MOM6
o2_by_location_mom6 <- skill_forecast %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(est, na.rm = TRUE)) %>%
  ungroup()

# Merge in situ and MOM6 mean O2 by location
o2_diff <- o2_by_location %>%
  rename(mean_o2_obs = mean_o2) %>%
  inner_join(o2_by_location_mom6 %>% rename(mean_o2_mom6 = mean_o2),
             by = c("longitude", "latitude")) %>%
  mutate(diff_o2 = mean_o2_obs - mean_o2_mom6)

# First plot: In situ O2
p1 <- ggplot(o2_by_location) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean In Situ O2 Values") +
  scale_color_viridis(option = "viridis", limits = c(0, max(o2_range)), name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Second plot: MOM6 O2
p2 <- ggplot(o2_by_location_mom6) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean MOM6 O2 Values") +
  scale_color_viridis(option = "viridis", limits = c(0, max(o2_range)), name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Third plot: Difference (MOM6 - In Situ)
p3 <- ggplot(o2_diff) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = diff_o2),
             size = 0.5, alpha = 0.7) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Difference: In Situ - MOM6", color = "ΔO2 (µmol/kg)") +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-diff_range, diff_range)
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10))

# Combine plots side by side with shared legend
p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

## Temporal trends

Next, lets see if the three scenarios align temporally (i.e., are they following roughly the same trend through time) starting with trends on an annual scale.

```{r}
#| echo: false
#| message: false
#| warning: false

# Combined original o2, nearest neighbor, and skill test analysis values into one dataset
skill_preds <- left_join(skill_preds, nn)

# add depth bins for anomaly calculation later
skill_preds <- skill_preds %>%
  mutate(depth_bin = cut(depth, breaks = seq(0, max(depth, na.rm=TRUE)+5, by=5),
                         include.lowest = TRUE, right = FALSE))
# add year month column for anomaly calculation later
skill_preds <- skill_preds %>%
  mutate(year_month = as.Date(paste(year, month, "01", sep = "-")))

# Reshape into long format
skill_preds_long <- skill_preds %>%
  select(year, month, o2, est, o2_nn_interp, depth, year_month) %>%
  pivot_longer(cols = c(o2, est, o2_nn_interp),
               names_to = "variable",
               values_to = "value")

# Summarize mean and 95% CI by year and variable
skill_preds_summary <- skill_preds_long %>%
  group_by(year, variable) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    se = sd(value, na.rm = TRUE) / sqrt(n()),   # standard error
    .groups = "drop"
  ) %>%
  mutate(
    lower = mean - 1.96 * se,
    upper = mean + 1.96 * se
  )

# Plot
ggplot(skill_preds_summary, aes(x = year, y = mean, color = variable)) +
  geom_line() +
  geom_point() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = variable), 
              alpha = 0.2, color = NA) +
  plot_theme +
  labs(x = "Year", y = "Oxygen (mean ± 95% CI)", 
       color = NULL, fill = NULL) +
  scale_color_manual(values = c("est" = "blue",
                                "o2_nn_interp" = "red",
                                "o2" = "black"),
                     labels = c("est" = "Indivero",
                                "o2_nn_interp" = "Nearest neighbor",
                                "o2" = "In situ")) +
  scale_fill_manual(values = c("est" = "blue",
                               "o2_nn_interp" = "red",
                               "o2" = "black"),
                    labels = c("est" = "Indivero",
                               "o2_nn_interp" = "Nearest neighbor",
                               "o2" = "In situ"))

```

Here we can see trend shape similarities between all three scenarios with the Indivero scenario deviating from the shape of the in situ trend only in 2011 and the nearest neighbor scenario resembling a flattened version of the in situ trend. Next let's try adjusting the scale to a year-month resolution.

```{r}
#| echo: false
#| message: false
#| warning: false

# Summarize mean and 95% CI by year and variable
skill_preds_summary <- skill_preds_long %>%
  group_by(year_month, variable) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    se = sd(value, na.rm = TRUE) / sqrt(n()),   # standard error
    .groups = "drop"
  ) %>%
  mutate(
    lower = mean - 1.96 * se,
    upper = mean + 1.96 * se
  )

# plot
ggplot(skill_preds_summary, aes(x = year_month, y = mean, color = variable)) +
  geom_line() +
  geom_point() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = variable), 
              alpha = 0.2, color = NA) +
  plot_theme +
  labs(x = "Year", y = "Oxygen (mean ± 95% CI)", 
       color = NULL, fill = NULL) +
  scale_color_manual(values = c("est" = "blue",
                                "o2_nn_interp" = "red",
                                "o2" = "black"),
                     labels = c("est" = "Indivero",
                                "o2_nn_interp" = "Nearest neighbor",
                                "o2" = "In situ")) +
  scale_fill_manual(values = c("est" = "blue",
                               "o2_nn_interp" = "red",
                               "o2" = "black"),
                    labels = c("est" = "Indivero",
                               "o2_nn_interp" = "Nearest neighbor",
                               "o2" = "In situ"))
```

Here we can see that both analyses generally follow the same temporal trend as the in situ scenario but with greater overall estimated values. The nearest neighbor scenario estimates increasing O2 values as months progress within each year which more closely resembles the trend of the in situ data.

## Anomalies

Now let's remove some of the mean biases from each scenario and look at variability through time by comparing dataset mean anomalies (i.e., not a climatological anomaly) on an annual and depth scale. Depths are grouped into 5 m depth bins.

```{r}
#| echo: false
#| message: false
#| warning: false

# remove rows with NAs in the nearest neighbor values to match sample size in all scenarios
skill_preds_anom <- skill_preds
skill_preds_anom <- skill_preds_anom[!is.na(skill_preds_anom$o2_nn_interp),]

# create anomaly dataset
skill_preds_anom <- skill_preds_anom %>%
  group_by(depth_bin) %>%
  mutate(o2_anom = o2 - mean(o2, na.rm = TRUE),
         est_anom = est - mean(est, na.rm = TRUE),
         o2_nn_interp_anom = o2_nn_interp - mean(o2_nn_interp, na.rm = TRUE)) %>%
  ungroup()

# Reshape into long format
anom_long <- skill_preds_anom %>%
  select(year, month, o2_anom, est_anom, o2_nn_interp_anom, depth, depth_bin, year_month) %>%
  pivot_longer(cols = c(o2_anom, est_anom, o2_nn_interp_anom),
               names_to = "variable",
               values_to = "value")

# Summarize mean and 95% CI by year and variable
anom_summary <- anom_long %>%
  group_by(year, variable) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    se = sd(value, na.rm = TRUE) / sqrt(n()),   # standard error
    .groups = "drop"
  ) %>%
  mutate(
    lower = mean - 1.96 * se,
    upper = mean + 1.96 * se
  )

# Plot
ggplot(anom_summary, aes(x = year, y = mean, color = variable)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = variable),
              alpha = 0.2, color = NA) +
  plot_theme +
  labs(x = "Year", y = "Oxygen anomaly (mean ± 95% CI)",
       color = NULL, fill = NULL) +
  scale_color_manual(values = c("est_anom" = "blue",
                                "o2_nn_interp_anom" = "red",
                                "o2_anom" = "black"),
                     labels = c("est_anom" = "Indivero",
                                "o2_nn_interp_anom" = "Nearest neighbor",
                                "o2_anom" = "In situ")) +
  scale_fill_manual(values = c("est_anom" = "blue",
                               "o2_nn_interp_anom" = "red",
                               "o2_anom" = "black"),
                    labels = c("est_anom" = "Indivero",
                               "o2_nn_interp_anom" = "Nearest neighbor",
                               "o2_anom" = "In situ"))

# # violin
# ggplot(anom_long, aes(x = factor(year), y = value, fill = variable)) +
#   geom_violin(alpha = 0.4, position = position_dodge(width = 0.8)) +
#   geom_boxplot(width = 0.1, position = position_dodge(width = 0.8),
#                outlier.shape = NA, alpha = 0.6) +
#   geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
#   plot_theme +
#   labs(x = "Year", y = "Oxygen anomaly", fill = NULL) +
#   scale_fill_manual(values = c("est_anom" = "blue",
#                                "o2_nn_interp_anom" = "red",
#                                "o2_anom" = "black"),
#                     labels = c("est_anom" = "Indivero",
#                                "o2_nn_interp_anom" = "Nearest neighbor",
#                                "o2_anom" = "In situ"))


```

And look at anomalies on a year-month scale.

```{r}
#| echo: false
#| message: false
#| warning: false

# Summarize mean and 95% CI by year and variable
anom_summary <- anom_long %>%
  group_by(year_month, variable) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    se = sd(value, na.rm = TRUE) / sqrt(n()),   # standard error
    .groups = "drop"
  ) %>%
  mutate(
    lower = mean - 1.96 * se,
    upper = mean + 1.96 * se
  )

# plot
ggplot(anom_summary, aes(x = year_month, y = mean, color = variable)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = variable), 
              alpha = 0.2, color = NA) +
  plot_theme +
  labs(x = "Year", y = "Oxygen anomaly (mean ± 95% CI)", 
       color = NULL, fill = NULL) +
  scale_color_manual(values = c("est_anom" = "blue",
                                "o2_nn_interp_anom" = "red",
                                "o2_anom" = "black"),
                     labels = c("est_anom" = "Indivero",
                                "o2_nn_interp_anom" = "Nearest neighbor",
                                "o2_anom" = "In situ")) +
  scale_fill_manual(values = c("est_anom" = "blue",
                               "o2_nn_interp_anom" = "red",
                               "o2_anom" = "black"),
                    labels = c("est_anom" = "Indivero",
                               "o2_nn_interp_anom" = "Nearest neighbor",
                               "o2_anom" = "In situ"))

# # violin
# ggplot(anom_long, aes(x = factor(year_month), y = value, fill = variable)) +
#   geom_violin(alpha = 0.4, position = position_dodge(width = 0.8)) +
#   geom_boxplot(width = 0.1, position = position_dodge(width = 0.8),
#                outlier.shape = NA, alpha = 0.6) +
#   geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
#   plot_theme +
#   labs(x = "Year", y = "Oxygen anomaly", fill = NULL) +
#   scale_fill_manual(values = c("est_anom" = "blue",
#                                "o2_nn_interp_anom" = "red",
#                                "o2_anom" = "black"),
#                     labels = c("est_anom" = "Skill test",
#                                "o2_nn_interp_anom" = "Nearest neighbor",
#                                "o2_anom" = "In situ"))
```

And finally look at anomalies spatially, comparing in situ to the Indivero method first.

```{r}
#| echo: false
#| message: false
#| warning: false

# calculate mean o2 by location- in situ
o2_by_location <- skill_preds_anom %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2_anom, na.rm = TRUE)) %>%
  ungroup()

# calculate mean o2 by location- MOM6
o2_by_location_mom6 <- skill_preds_anom %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(est_anom, na.rm = TRUE)) %>%
  ungroup()

# calculate mean o2 by location- nearest neighbor
o2_by_location_nn <- skill_preds_anom %>%
  group_by(longitude, latitude) %>%
  dplyr::summarise(mean_o2 = mean(o2_nn_interp_anom, na.rm = TRUE)) %>%
  ungroup()

# Merge in situ and MOM6 mean O2 by location
o2_diff <- o2_by_location %>%
  rename(mean_o2_obs = mean_o2) %>%
  inner_join(o2_by_location_mom6 %>% rename(mean_o2_mom6 = mean_o2),
             by = c("longitude", "latitude")) %>%
  mutate(diff_o2 = mean_o2_obs - mean_o2_mom6)

# Merge in situ and nearest neighbor mean O2 by location
o2_diff_nn <- o2_by_location %>%
  rename(mean_o2_obs = mean_o2) %>%
  inner_join(o2_by_location_nn %>% rename(mean_o2_mom6 = mean_o2),
             by = c("longitude", "latitude")) %>%
  mutate(diff_o2 = mean_o2_obs - mean_o2_mom6)

# Find the global O2 range across both datasets
o2_range_anom <- range(c(o2_by_location$mean_o2, o2_by_location_mom6$mean_o2, o2_by_location_nn$mean_o2), na.rm = TRUE)

# Find symmetric range for difference plot color scale
diff_range_anom <- max(abs(c(o2_diff$diff_o2, o2_diff_nn$diff_o2)), na.rm = TRUE)

# First plot: In situ O2
p1 <- ggplot(o2_by_location) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean In Situ O2 Values") +
  scale_color_viridis(option = "viridis", limits = o2_range_anom, name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Second plot: MOM6 O2
p2 <- ggplot(o2_by_location_mom6) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean MOM6 O2 Values") +
  scale_color_viridis(option = "viridis", limits = o2_range_anom, name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Third plot: Difference (MOM6 - In Situ)
p3 <- ggplot(o2_diff) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = diff_o2),
             size = 0.5, alpha = 0.7) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Difference: In Situ - MOM6", color = "ΔO2 (µmol/kg)") +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-diff_range_anom, diff_range_anom)
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10))

# Combine plots side by side with shared legend
p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

And then comparing the in situ to the nearest neighbor method.

```{r}
#| echo: false
#| message: false
#| warning: false

# Second plot: MOM6 O2
p2 <- ggplot(o2_by_location_nn) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = mean_o2),
             size = 0.5,
             alpha = 0.6) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Mean MOM6 O2 Values") +
  scale_color_viridis(option = "viridis", limits = o2_range_anom, name = "O2 (µmol/kg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 10))

# Third plot: Difference (MOM6 - In Situ)
p3 <- ggplot(o2_diff_nn) +
  geom_sf(data = namerica_state_wgs_df,
          fill = NA, color = "grey10", size = 0.2) +
  geom_point(aes(x = longitude, y = latitude, color = diff_o2),
             size = 0.5, alpha = 0.7) +
  ylab("Latitude") +
  xlab("Longitude") +
  xlim(min(nn$longitude, na.rm = TRUE), max(nn$longitude, na.rm = TRUE)) +
  ylim(min(nn$latitude, na.rm = TRUE), max(nn$latitude, na.rm = TRUE)) +
  labs(title = "Difference: In Situ - NN", color = "ΔO2 (µmol/kg)") +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-diff_range_anom, diff_range_anom)
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 10))

# Combine plots side by side with shared legend
p1 + p2 + p3 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

## Confusion matrix 

Thinking about this from a slightly different angle, let's simplify it to a binary scenario and calculate some evaluation metrics based on a confusion matrix. In this example we will label oxygen values less than hakes oxygen preference boundary (i.e., 119 µm/kg) as 1, and oxygen values greater than the preference boundary as 0.

### Indivero method

We'll start by comparing in situ to the Indivero method first.

```{r}
#| echo: false
#| message: false
#| warning: false

skill_preds_anom$o2_binary <- 0
skill_preds_anom$o2_binary[skill_preds_anom$o2 <= 119] <- 1

skill_preds_anom$est_binary <- 0
skill_preds_anom$est_binary[skill_preds_anom$est <= 119] <- 1

# Confusion matrix components
TP <- as.numeric(sum(skill_preds_anom$o2_binary == 1 & skill_preds_anom$est_binary == 1))
TN <- as.numeric(sum(skill_preds_anom$o2_binary == 0 & skill_preds_anom$est_binary == 0))
FP <- as.numeric(sum(skill_preds_anom$o2_binary == 0 & skill_preds_anom$est_binary == 1))
FN <- as.numeric(sum(skill_preds_anom$o2_binary == 1 & skill_preds_anom$est_binary == 0))

confusion_matrix <- matrix(
  c(TP, FP, FN, TN),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    "Observed" = c("Positive", "Negative"),
    "Predicted" = c("Positive", "Negative")
  )
)

confusion_matrix

```

From the confusion matrix we can see relatively low false positive rates (390) and higher false negative rates (15843) indicating that the model is overpredicting oxygen values, which mirrors what we saw in the analysis above.

```{r}
#| echo: false
#| message: false
#| warning: false

# calculate true skill statistic (TSS) which is better for imbalanced classes
TPR <- TP / (TP + FN)
TNR <- TN / (TN + FP)
TSS <- TPR + TNR - 1
print("True Skill Statistic")
TSS

# calculate Matthews Correlation Coefficient (MCC) which is considered the most \
# informative metric to evaluate a confusion matrix
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP)*(TP + FN)*(TN + FP)*(TN + FN))
print("Matthews Correlation Coefficient")
MCC
```

Finally, we can calculate True Skill Statistic and Matthews Correlation Coefficent evaluation metrics that account for class imbalances. Both range from -1 (inverse prediction) to 1 (perfect prediction) with 0 indicating classification no better than random. There isn’t a single universal cutoff for values that are considered acceptable, but in practice, TSS and MCC values of 0-0.2 are considered very weak, 0.2-0.4 have weak to moderate skill, 0.4-0.6 have moderate to good skill, 0.6-0.8 have strong skill, and 0.8 to 1.0 have near perfect skill. Both our TSS and MCC values fall within the strong skill range suggesting a strong, reliable classification model.

### Nearest neighbor method

Then we compare in situ to the nearest neighbor method.

```{r}
#| echo: false
#| message: false
#| warning: false

skill_preds_anom$o2_binary <- 0
skill_preds_anom$o2_binary[skill_preds_anom$o2 <= 119] <- 1

skill_preds_anom$o2_nn_interp_binary <- 0
skill_preds_anom$o2_nn_interp_binary[skill_preds_anom$o2_nn_interp <= 119] <- 1

# Confusion matrix components
TP <- as.numeric(sum(skill_preds_anom$o2_binary == 1 & skill_preds_anom$o2_nn_interp_binary == 1))
TN <- as.numeric(sum(skill_preds_anom$o2_binary == 0 & skill_preds_anom$o2_nn_interp_binary == 0))
FP <- as.numeric(sum(skill_preds_anom$o2_binary == 0 & skill_preds_anom$o2_nn_interp_binary == 1))
FN <- as.numeric(sum(skill_preds_anom$o2_binary == 1 & skill_preds_anom$o2_nn_interp_binary == 0))

confusion_matrix <- matrix(
  c(TP, FP, FN, TN),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    "Observed" = c("Positive", "Negative"),
    "Predicted" = c("Positive", "Negative")
  )
)

confusion_matrix

```

From the confusion matrix we can see relatively low false positive rates (1369) and higher false negative rates (12011) indicating that the model is overpredicting oxygen values, which mirrors what we saw in the analysis above.

```{r}
#| echo: false
#| message: false
#| warning: false

# calculate true skill statistic (TSS) which is better for imbalanced classes
TPR <- TP / (TP + FN)
TNR <- TN / (TN + FP)
TSS <- TPR + TNR - 1
print("True Skill Statistic")
TSS

# calculate Matthews Correlation Coefficient (MCC) which is considered the most \
# informative metric to evaluate a confusion matrix
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP)*(TP + FN)*(TN + FP)*(TN + FN))
print("Matthews Correlation Coefficient")
MCC
```

And again, both our TSS and MCC values fall within the strong skill range suggesting a strong, reliable classification model.

## Conclusions

Our evaluation indicates that the raw MOM6 output (i.e., the nearest neighbor method) is suitable for inclusion in the forecasting model. The nearest neighbor method consistently outperformed the Indivero method across all RMSE comparisons, reproduced temporal trends observed in the in situ data, and demonstrated strong predictive skill in the binary framework (TSS and MCC). RMSE values at the focal depths of the forecasting analysis were within an acceptable range (15–20; Indivero, pers. comm., 2025-08-22). While MOM6 tended to overestimate oxygen concentrations in shallower waters, these biases occurred primarily above the oxygen preference boundary, as reflected by the lower false-positive rate (n = 1369), and are unlikely to influence hake distribution. More importantly, however, MOM6 also overestimated oxygen at depths of \~100–200 m, where oxygen values may be lower than the preference boundary value, which could lead to localized overprediction of hake biomass in the forecast model.

## References
